Phylogeographic inference with microtellite and SNPs

### 2019-07-02 code/rmarkdown/article.Rmd
kfinder plots

## 2.3 Population structure
We conducted model-based genetic structure analyses in STRUCTURE v2.3.4 (Pritchard et al. 2000). For each dataset, we performed 10 replicate runs from K = 1 to K = 8, since the best number of clusters fall within this range for both species (Gutiérrez-Rodríguez et al. 2017; Sánchez-Montes et al. 2019). We used an admixture model with correlated allele frequencies (Falush et al. 2003) and no prior information of sample origin. For the microtellite data, we used the same run lengths as in their original publications: 5*10^5 burnin followed by 1*10^6 iterations. For the SNP datasets, the run lengths were shorter, a burnin of 3*10^4 steps followed by 1 * 10^4 iterations, as for SNP data chain often converge fast (Pritchard, google groups...?). For the SNP data, we estimated lambda for K = 1, by averaging lambda estimates across 3 replicate runs. The values of lambda (0.60 for *H. molleri* and 0.69 for *P. cultripes*) were then used across all runs for the SNP data, whereas for microsatellites lambda was fixed to 1. These lower values of lambda can improve the modeling of correlated allele frequencies when using SNPs, where often the data is skewed toward rare alleles (Falush et al. 2003). STRUCTURE was run in parallel in 8 cores using Structure_threader (Pina-Martins et al. 2017), recording steps to log files every 50 and 5000 iterations for the SNP and microsatellite data, respectively. We run STRUCTURE analysis with random subsets of loci (n = 200, 500, 1000, 3000, 5000, 10000 and 20000) from the original matrix of SNPs for *P. cultripes* and the same STRUCTURE parameters.
The convergence between the 10 replicate runs for each K was evaluated in R using Gelman and Rubin's convergence diagnostic (GR diagnostic; Gelman and Rubin 1992), with function *gelman.diag* in package *coda* (Plummer et al. 2006).

@@

## 3.1 Model-based clustering
The convergence of the chains as measured by GR diagnostic varied between speciesa and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).

>>>>

## 2.3 Population structure
We conducted model-based genetic structure analyses in STRUCTURE v2.3.4 (Pritchard et al. 2000). For each dataset, we performed 10 replicate runs from K = 1 to K = 8, since the best number of clusters fall within this range for both species (Gutiérrez-Rodríguez et al. 2017; Sánchez-Montes et al. 2019). We used an admixture model with correlated allele frequencies (Falush et al. 2003) and no prior information of sample origin. For the microtellite data, we used the same run lengths as in their original publications: 5*10^5 burnin followed by 1*10^6 iterations. For the SNP datasets, the run lengths were shorter, a burnin of 3*10^4 steps followed by 1 * 10^4 iterations, as for SNP data chain often converge fast (Pritchard, google groups...?). For the SNP data, we estimated lambda for K = 1, by averaging lambda estimates across 3 replicate runs. The values of lambda (0.60 for *H. molleri* and 0.69 for *P. cultripes*) were then used across all runs for the SNP data, whereas for microsatellites lambda was fixed to 1. These lower values of lambda can improve the modeling of correlated allele frequencies when using SNPs, where often the data is skewed toward rare alleles (Falush et al. 2003). STRUCTURE was run in parallel in 8 cores using Structure_threader (Pina-Martins et al. 2017), recording steps to log files every 50 and 5000 iterations for the SNP and microsatellite data, respectively. We run STRUCTURE analysis with random subsets of loci (n = 200, 500, 1000, 3000, 5000, 10000 and 20000) from the original matrix of SNPs for *P. cultripes* and the same STRUCTURE parameters. This approximation allowed to assess the effect of the number of loci in the patterns of genetic structure.
The convergence between the 10 replicate runs for each K was evaluated in R using Gelman and Rubin's convergence diagnostic (GR diagnostic; Gelman and Rubin 1992), with function *gelman.diag* in package *coda* (Plummer et al. 2006).
A central question in population genetics is determining the optimal number of populations subjacent to the genetic structure (Allendorf and Luikart 2007). However, hierarchical population structure (Evanno et al. 2005), unbalanced sampling and number markers (Wang 2017) can lead to the incorrect inference of number of populations. We used KFinder (Wang 2019) to compare the best number of populations for each dataset through 3 approaches: (1) Pr[X|K], the probability of data X given K populations (Pritchard et al. 2000) and, (2) ΔK, the most popular method and which considers the rate of change in the log probability of data between successive K (Evanno et al. 2005) and 3) PI, parsimony index, a newly proposed metric which favors K's which yield most consistent and with minimal average individual admixture, and which claims to be more consistent across wider demographic scenarios (Wang 2019).

@@

## 3.1 Model-based clustering
Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
